Verifying Spatial Properties of Stencil Computations

### 2016-11-16 paper.tex
Replace use of refs with Cref

@@

In this paper, we make the following contributions:
%
\begin{itemize}
\item We introduce a specification language for stencils
  that captures common forms of data access patterns
  (\S~\ref{sec:lang});

\item We detail checking and inference algorithms for our
  specifications (\S~\ref{sec:analysis}), derived from a
  denotational model for the specificaiton language
  (\S~\ref{sec:semantics});

@@

\item We report on a quantitative study of stencil computations on a
  corpus of numerical Fortran program
  (\S~\ref{sec:evaluation}), totalling one million lines of code. Our tool
  identifies and infers specifications for 35,000 stencil computations in the
  corpus. Approximately 10\% of the stencils we found are non-trivial,
  corresponding to code which is a possible source of errors. This validates
  our hypothesis that the majority of stencil computations have a regular
  shape, and validates the design of our language in its capability to capture
  many core patterns.

@@

\newcommand{\domainVal}{\mathbb{Z}_\infty}
\section{Stencil specifications}

Our specification language is based on the hypothesis
that most forms of array access in numerical code have
a fixed, statically-determined access pattern. For example, the
\emph{five-point stencil} on a two-dimensional array reads from array
indices $(i, j)$, $(i-1, j)$, $(i+1, j)$, $(i, j-1)$, and $(i, j+1)$
for all $i, j$ within the inner boundary of the array (to avoid
out-of-bounds access at the edges). We revisit this hypothesis
in \S~\ref{sec:evaluation} with the inference of
such regular stencil patterns on our corpus of numerical programs.

We give various definitions used in the paper then
\S~\ref{sec:syntax} outlines the syntax of stencil specifications
with an informal introduction. \S~\ref{sec:semantics}
gives a semantic model via a set-based interpretation which makes
concrete the meaning of specifications. \S~\ref{sec:eqs} then provides
a syntactic equational theory for specifications which is
sound with respect to the model.
%sub-specification for which the
%equality and sub-specification relation are proven sound.

@@

Now we informally describe what it means for a stencil kernel to be consistent
with a specification. This will be formalised in \S~\ref{sec:semantics} and
algorithmically described in  and \S~\ref{sec:analysis}.
Specifications are associated to a stencil kernel and are declared for one or
more array variables $\bar{v}$ within that kernel.  A stencil specification
defines a \emph{region} associated to an array variable $a$ against which a
stencil kernel should be \emph{consistent}. The region defined by a
specification is an $n$-dimensional rectilinear shape in a discrete space
crossing the origin with respect to the induction variable. Consistency is then
a two way requirement. All array subscripts on $a$
contributing to the kernel should be consistent with the region. Conversely,
all parts of the region should have an index corresponding it.

@@

\subsection{Specification syntax}
\label{sec:syntax}

Figure~\ref{fig:syntax} gives the syntax of stencil specifications, which is
detailed below. The entry point is the \nonterm{specification} production which
splits into either a \emph{region declaration} or a \emph{specification
declaration}.

@@

Region specifications can be assigned to region variables via
region declarations. For example, the shape of a
``\emph{Roberts cross}'' edge-detection convolution~\cite{davis1975survey}
can be stated:
%%
\begin{minted}{fortran}
!= region :: r1 = forward(depth=1, dim=1)
!= region :: r2 = forward(depth=1, dim=2)
!= region :: robertsCross = r1*r2
!= stencil robertsCross :: a
\end{minted}
This is useful for common stencil patterns, such as the five-point
pattern, as the region can be defined once and reused.
%%
\paragraph{Modifiers}
%%
Region specifications can be modified
by \emph{approximation} and \emph{multiplicity} information
(in \textit{spec} in Fig.~\ref{fig:syntax}).

@@

We define a simple set-theoretic denotational model of the semantics
of our specification language. This has a number of purposes:
(1) it serves to explain the meaning of our specifications
(2) it is used in the inference and checking algorithms
(Section~\ref{sec:analysis}); (3) justifies an equational theory for
specifications in the next section; and (4) can be used to guide
correct implementations. The model domain is over sets of vectors
which we call \emph{index schemes}.

\begin{definition}
  An \emph{index scheme} is a vector of size $n$ (an $n$-vector) with
  values drawn from $\domainVal = \mathbb{Z} \cup \{\infty\}$.  Values
  in $\mathbb{Z}$ represent the offsets of neighbourhood indices
  (\ie{}, the constant integer $c$ in an indexing expression of the
  form $i \pm c$, Def.~\ref{def:neighbour}) and $\infty$ is a wild-card
  representing both constant indices and unspecified indices.
  For an index scheme $u$, $u_i$ denotes the $i^{th}$ element of $u$. We
  write $(\domainVal)^n$ for the set of all $n$-dimensional schemes.
\end{definition}

Indexing schemes abstract array subscripts.
Our model represents the meaning of stencil specifications
as indexing schemes, which explain the range of possible
indexing behaviours which are consistent with the specification.
The relationship between source language
terms and indexing terms
 is given precisely by the following function \textsf{schematic},
 which forms part of checking and inference (\S\ref{sec:analysis}).

@@

The model is defined by the interpretation function $\interp{-}_n$
(where $n$ is the maximum dimensionality of the specification being
modelled) mapping closed\footnote{That is, we assume there are no
  occurrences of \textit{rvar} in a specification being modelled.  Any
  \emph{open} specification containing region variables can be made
  closed by straightforward syntactic substitution with a (closed)
  \textit{region}.} specifications to sets of
$n$-dimensional indexing schemes. The interpretation is overloaded on
\emph{regions} in Figure~\ref{fig:region-model} and on the top-level
of a specification \textit{spec} in Figure~\ref{fig:spatial-model}.

@@

\noindent
The first four equations of Fig.~\ref{fig:region-model}
give the model of the region constants as sets of indexing
schemes. We illustrate with an example.

\begin{example}
For a $2$-dimensional stencil computation, then
\begin{align*}
& \interp{\stenCenS{2}{1}}_2 \\[-0.1em]
& \qquad = \; \{i\singleEntry{1}{2} \,|\, i \in \{\text{-}2,..., 2\}\!\setminus\!0 \}
  \cup  \{ \zeroEntry{1}{2} \,|\, \textit{true} \} \\[-0.2em]
& \qquad = \; \{-2\singleEntry{1}{2}, \, -\singleEntry{1}{2}, \,
  \singleEntry{1}{2}, \, 2\singleEntry{1}{2}\} \cup
  \{\zeroEntry{1}{2}\} \\[-0.2em]
& \qquad = \; \{\vtwoh{-2}{\infty}, \vtwoh{-1}{\infty}, \vtwoh{0}{\infty},
\vtwoh{1}{\infty}, \vtwoh{2}{\infty}\}
\end{align*}
Note the absorbing behaviour of $\infty$ with respect to
multiplication.
\end{example}
\noindent
The final two equations of Fig.~\ref{fig:region-model} give the models
of $\term{+}$ and $\term{*}$. For $\term{+}$ this is the
straightforward union of the models of subterms. The model of
$\term{*}$ takes the tensor $\otimes$ of subterm models. The tensor is
more involved; we give some informal intuition first.

@@

Regions are modelled as sets of index
schemes, that is
$\interp{\textit{region}}_n \in \mathcal{P}(\domainVal{}^n)$.
Figure~\ref{fig:spatial-model} defines the top-level model of the
\textit{spec} syntax, which includes the multiplicity and
approximation modifiers. The domain of this interpretation
is $\textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$
where $\textsf{Mult}$ and $\textsf{Approx}$ are labelled variants
labelling our set-based model with additional information.

\begin{definition} \textsf{Mult}
and \textsf{Approx} are parametric labelled variant types
with injection function given by their definition:
%
\begin{align*}
\textsf{Mult} \, a & = \textsf{mult} \, a \,\mid\, \textsf{only} \, a \\
\textsf{Approx} \, a & = \textsf{exact} \, a \,\mid\, \textsf{lower} \, a \,\mid\,
\textsf{upper} \, a \,\mid\, \textsf{both} \, a \, a
\end{align*}
\eg{}, $\mathsf{lower}$ is an injection
$\mathsf{lower} : a \rightarrow \mathsf{Approx} \, a$ etc. The
$\textsf{Mult}$ type corresponds to the presence or absence of the
\term{readOnce} modifier as shown in Figure~\ref{fig:spatial-model}.
The $\textsf{Approx}$ type corresponds to the presence or absence of
the spatial approximation modifier, with \textsf{exact} when there is
no such modifier and \textsf{lower} and \textsf{upper} for
\term{atLeast} and \term{atMost}. Later, it will be useful to model
pairs of lower and upper bounds provided by \textsf{both} but which is
not used
here.% These are used for labelled models, and other components later in.
\label{def:mult-and-approx}
\end{definition}

@@

The equality relation $\equiv$ is defined over \emph{region}
syntax which gives the following axioms:
%
\begin{itemize}
\item $\term{+}$ is idempotent, commutative, and associative;
\item $\term{*}$ is commutative, associative, and distributes over
$\term{+}$, \ie{},  $R \; \texttt{*} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{+} \; (R
                       \; \texttt{*} \; T)$.
Distributivity is
used later to give a normal form for specifications
akin to Disjunctive Normal Form (\S~\ref{sec:inf-step4}).

@@

The equational theory provides not only an axiomatic semantics
for the language, which the user can use to rewrite their
specification, but it is also used extensively in the inference
procedure (\S~\ref{subsec:inference}) to simplify specifications.

@@

Figure~\ref{fig:inequations} defines a notion of sub-specifications
via the relation $<:^r$ on \textit{region}
syntax and then $<:$ on \textit{spec} syntax inductively.
Both relations are reflexive and transitive (we omit
these equations). The $<:^r$ relation is congruent
with respect to \term{+} and \term{*} $<:$ is congruent
with repect to \term{readOnce}, the equations of which we omit for brevity.

@@

\begin{theorem}[Approximation soundness]
Code consistent with a specification $S$
 is consistent with a more general specification $T$,
\ie{}, for all models $M$ of a piece of code:
\begin{equation*}
S <: T \; \wedge (\consA{M}{\interp{S}_n}) \;
\Rightarrow \; (\consA{M}{\interp{T}_n})
\end{equation*}
where $\consAName$ implements consistency in checking (\S~\ref{subsec:checking}).
\end{theorem}

@@

\noindent
We outline here the procedures for checking conformance
of source code against specifications (\S~\ref{subsec:checking})
and for infering specifications from code (\S~\ref{subsec:inference}).
Both rely on a program analysis that converts array indexing
expressions into sets of index schemes. We outline this analysis
first (\S~\ref{subsec:analysis}). Note that the analysis procedure
can be made arbitrarily more complex and wide-ranging independent
of the checking and inference procedures. At the moment, the analysis
is largely \emph{syntactic}, with only a small amount of
semantic interpretation of the code.

@@

The core analysis, captured by the function $\textit{analyse}$ below, then proceeds as follows.
For each assignment statement \textit{assg} within a loop whose left-hand side is an array
subscript on neighbourhood indices, a finite map is computed mapping
array variables to \emph{multisets} of indexing schemes:
%
\begin{align*}
\hspace{-0.5em}
\begin{array}{lr}
\textit{analyse}(a(\overline{e_1}) = e_2)
 := \qquad\qquad\qquad & \textit{where} \; \neigh(\overline{e_1}) \\[0.3em]
\multicolumn{2}{l}{\qquad \bigcup\{a' \mapsto \{
\textsf{schematic}(\bar{e})\} \mid a'(\bar{e}) \leftarrow \mathsf{flowsTo}(e_2)\}}
\end{array}
\end{align*}
%
The predicate $\neigh(\bar{e_1})$ classifies indices where each
$e \in \bar{e_1}$ is either a
neighbourhood index (based on
the induction variables of the loop, implicit here)
or a constant index.\footnote{If all indices $e \in \bar{e}$ are constants, then
  $\neigh(\bar{e})$ does not hold. Instead $a(\bar{e}$) is
  considered a constant by our analysis, rather
than an array subscript.}% contributing to a stencil.}
Thus, we focus on assignments to an array subscript where the LHS
indices $\overline{e_1}$ are neighbourhood indidces.  For
all array subscripts that flow to the right-hand side of this
statement, a finite map is constructed, mapping each array variable
to a multiset of indexing schemes for its subscripts, computed
with \textsf{schematic} (Def.~\ref{def:schematic}).
Note \textsf{schematic} is undefined ($\bot$) if
$\bar{e}$ contains relative indices which are not neighbourhood
indices. The \textit{analysis} function is used by the
\textit{modelise} function which filters out those arrays
which contain undefined indexing schemes and adds multiplicity
information:
%
\begin{align*}
& \mathit{modelise} (\textit{assg})
= \textit{let} \;\; M = \mathit{analyse}(\textit{assg}) (a) \;\; \textit{in}
\\
& a \mapsto
\begin{cases}
\bot & \bot \in M \\
\textsf{only}(|M|) & \forall u \in M . \, \#(M, u) = 1 \\
\textsf{mult}(|M|) & \forall u \in M . \, \#(M, u) = n
\end{cases}
\end{align*}
where $\#(M, u)$ denotes the multiplicity of an element $u$ of the
multiset $M$ and $|M|$ denotes the set of elements in the multiset
$M$.
Thus, the analysis maps source code to index schemes, which is then
augmented with multiplicity information by $\mathit{modelise}$,
\ie{} values in $\mathsf{Mult}(\mathcal{P}((\domainVal{})^n))$.

@@

Checking verifies the access pattern of a stencil computation in the
source language against its associated specifications. The model of
(\S~\ref{sec:semantics}) guides this process.  Checking
proceeds by generating a model from a specification and generating
a model from the source code, comparing them for
\emph{consistency} base on index scheme
\emph{generalisation} (Def.~\ref{def:generalisation}).

@@

A stencil kernel
 $v(\overline{e_1}) = e_2$
conforms to $S$ if the predicate $\textbf{check}_n(S, v(\overline{e_1}) = e_2, a)$ holds,
defined:
\begin{align*}
& \textbf{check}_n(S, \textit{assg}, a) =
  \consSub{n}{\textit{modelise}(\textit{assg})(a)}{\interp{S}_n}
&
\end{align*}
Thus, checking applies \textit{modelise}
(\S~\ref{subsec:analysis}) to extract index schemes from source code,
augmented with multiplicitly for the whole set of schemes.
This is compared with the model of the specification $\interp{S}_n$ by
the consistency predicate $\consAName$.
\end{definition}

@@

\begin{definition}
Consistency relates models of source code to models
of specifications (which are bounded) via predicates:
\begin{align*}
\consName_n & \subseteq \mathcal{P}(\domainVal{}^n)
\times \textsf{Approx}(\mathcal{P}(\domainVal{}^n)) \\
\consAName_n & \subseteq \textsf{Mult}(\mathcal{P}(\domainVal{}^n))
\times \textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))
\end{align*}
%
where $\consName$ relates models without multiplicity.
Figure~\ref{fig:consistency} defines the two predicates, where for
brevity we make implicit the subscript $n$ defining the
dimensionality.  We denote sets of indexing schemes
$\mathcal{P}((\domainVal{})^n)$
from source code model by $M$ and from a specification model by $M$.

@@

To aid adoption of our approach to legacy code, we provide
an inference procedure for generating specifications from code, which
are inserted automatically as comments.
We illustrate inference using the
 five-point stencil of
 Example~\ref{exm:checking}.

@@

The first step performs the
analysis of \S~\ref{subsec:analysis} to produce a
model of the code in terms of indexing schemes.
The \emph{analyse} and \emph{modelise} procedures before yielded the finite map:
%
\begin{align*}
\texttt{a} \mapsto \mathsf{only}(\{\vtwoh{-1}{0}, \vtwoh{1}{0},
          \vtwoh{0}{-1}, \vtwoh{0}{1}, \vtwoh{0}{0}\})
\end{align*}
%

@@

\noindent
Each indexing scheme $u \in U(a)$ is converted to a \emph{unit} span
$u \mapsto [u \; u]$, defining the lower and upper bound of an
$n$-dimensional cube of side length $1$. For the five-point stencil
example there are five unit spans, illustrated in
Figure~\ref{fig:inference-steps-informal}(a).

@@

\begin{definition}\label{def:span-coalesc}
  Span coalescing $\bullet$ is partial, defined for spans contiguous in the
  first dimension:
%%
\begin{align*}
\vect{x} \bullet \vect{y}
= (\vect{x}^L, \vect{y}^U) \;\;\;\, \textit{if} \;\;\; \vect{x}^U_1 + 1 = \vect{y}^L_1 \; \wedge \;
\vect{x}_{2:n} = \vect{y}_{2:n} %\\
%\bot  & \textit{otherwise}
%\end{cases}
% $m$-dimensions into coalesced spans which vary in
% $m+1$-dimensions.
\end{align*}
%%
\end{definition}
\noindent
To coalesce regions which are contiguous in other dimensions
other than the first, we generate all rotations of the spans in
$n$-dimensional space by permutation on spans, and fold together contiguous
regions using $\bullet$ (above). This
generates all possible ways of coalescing regions. For our example,
the unit spans in Figure~\ref{fig:inference-steps-informal}(a)
are coalesced into the contiguous spans in
Figure~\ref{fig:inference-steps-informal}(b).
This procedure is iterated until a fixed point is reached (in one step
in this example).
 Then any spans that are completely contained by any other span are deleted,
leaving a minimal set of spans (which may overlap, but none of which
fully contains another), Figure~\ref{fig:inference-steps-informal}(c).
The final spans are used to synthesise concrete stencil specification
syntax (\S~\ref{sec:inf-step4}), where
each span becomes a product \texttt{*} of regions which are
combined by the sum \texttt{+}.

@@

\begin{definition} The $\mathsf{toSpec}$ constructor is defined
in Figure~\ref{fig:to-spec}, where the lower and upper bounds
determine region constants; $\mathsf{toSpec} \, d \, (l, u)$ is always defined
 if $l$ less than or equal to $u$.

@@

\paragraph{Hypothesis 1: Stencil computations
(as in Definition~\ref{def:stencil}) are a common programming pattern
in numerical code}

>>>>

\usepackage{cleveref}
\crefname{section}{\S}{\S\S}
\Crefname{section}{\S}{\S\S}
\crefname{subsection}{\S\S}{\S\S\S}
\Crefname{subsection}{\S\S}{\S\S\S}
\crefname{figure}{fig.}{}
\Crefname{figure}{Fig.}{}
\crefname{definition}{def.}{}
\Crefname{definition}{Def.}{}

@@

In this paper, we make the following contributions:
%
\begin{itemize}
\item We introduce a specification language for stencils
  that captures common forms of data access patterns
  (\cref{sec:lang});

\item We detail checking and inference algorithms for our
  specifications (\cref{sec:analysis}), derived from a
  denotational model for the specificaiton language
  (\cref{sec:semantics});

@@

\item We report on a quantitative study of stencil computations on a
  corpus of numerical Fortran program
  (\cref{sec:evaluation}), totalling one million lines of code. Our tool
  identifies and infers specifications for 35,000 stencil computations in the
  corpus. Approximately 10\% of the stencils we found are non-trivial,
  corresponding to code which is a possible source of errors. This validates
  our hypothesis that the majority of stencil computations have a regular
  shape, and validates the design of our language in its capability to capture
  many core patterns.

@@

\newcommand{\domainVal}{\mathbb{Z}_\infty}
\section{Stencil specifications}\label{sec:lang}

Our specification language is based on the hypothesis
that most forms of array access in numerical code have
a fixed, statically-determined access pattern. For example, the
\emph{five-point stencil} on a two-dimensional array reads from array
indices $(i, j)$, $(i-1, j)$, $(i+1, j)$, $(i, j-1)$, and $(i, j+1)$
for all $i, j$ within the inner boundary of the array (to avoid
out-of-bounds access at the edges). We revisit this hypothesis
in \cref{sec:evaluation} with the inference of
such regular stencil patterns on our corpus of numerical programs.

We give various definitions used in the paper then
\cref{subsec:syntax} outlines the syntax of stencil specifications
with an informal introduction. \cref{sec:semantics}
gives a semantic model via a set-based interpretation which makes
concrete the meaning of specifications. \cref{sec:eqs} then provides
a syntactic equational theory for specifications which is
sound with respect to the model.
%sub-specification for which the
%equality and sub-specification relation are proven sound.

@@

Now we informally describe what it means for a stencil kernel to be consistent
with a specification. This will be formalised in \Cref{sec:semantics} and
algorithmically described in  and \Cref{sec:analysis}.
Specifications are associated to a stencil kernel and are declared for one or
more array variables $\bar{v}$ within that kernel.  A stencil specification
defines a \emph{region} associated to an array variable $a$ against which a
stencil kernel should be \emph{consistent}. The region defined by a
specification is an $n$-dimensional rectilinear shape in a discrete space
crossing the origin with respect to the induction variable. Consistency is then
a two way requirement. All array subscripts on $a$
contributing to the kernel should be consistent with the region. Conversely,
all parts of the region should have an index corresponding it.

@@

\subsection{Specification syntax}
\label{subsec:syntax}

\Cref{fig:syntax} gives the syntax of stencil specifications, which is
detailed below. The entry point is the \nonterm{specification} production which
splits into either a \emph{region declaration} or a \emph{specification
declaration}.

@@

Region specifications can be assigned to region variables via
region declarations. For example, the shape of a
``\emph{Roberts cross}'' edge-detection convolution~\cite{davis1975survey}
can be stated:
%%
\begin{minted}{fortran}
!= region :: r1 = forward(depth=1, dim=1)
!= region :: r2 = forward(depth=1, dim=2)
!= region :: robertsCross = r1*r2
!= stencil robertsCross :: a
\end{minted}
This is useful for common stencil patterns, such as the five-point
pattern, as the region can be defined once and reused.
%%
\paragraph{Modifiers}
%%
Region specifications can be modified
by \emph{approximation} and \emph{multiplicity} information
(in \textit{spec} in \Cref{fig:syntax}).

@@

We define a simple set-theoretic denotational model of the semantics
of our specification language. This has a number of purposes:
(1) it serves to explain the meaning of our specifications
(2) it is used in the inference and checking algorithms
(\Cref{sec:analysis}); (3) justifies an equational theory for
specifications in the next section; and (4) can be used to guide
correct implementations. The model domain is over sets of vectors
which we call \emph{index schemes}.

\begin{definition}
  An \emph{index scheme} is a vector of size $n$ (an $n$-vector) with
  values drawn from $\domainVal = \mathbb{Z} \cup \{\infty\}$.  Values
  in $\mathbb{Z}$ represent the offsets of neighbourhood indices
  (\ie{}, the constant integer $c$ in an indexing expression of the
  form $i \pm c$, \Cref{def:neighbour}) and $\infty$ is a wild-card
  representing both constant indices and unspecified indices.
  For an index scheme $u$, $u_i$ denotes the $i^{th}$ element of $u$. We
  write $(\domainVal)^n$ for the set of all $n$-dimensional schemes.
\end{definition}

Indexing schemes abstract array subscripts.
Our model represents the meaning of stencil specifications
as indexing schemes, which explain the range of possible
indexing behaviours which are consistent with the specification.
The relationship between source language
terms and indexing terms
 is given precisely by the following function \textsf{schematic},
 which forms part of checking and inference (\Cref{sec:analysis}).

@@

The model is defined by the interpretation function $\interp{-}_n$
(where $n$ is the maximum dimensionality of the specification being
modelled) mapping closed\footnote{That is, we assume there are no
  occurrences of \textit{rvar} in a specification being modelled.  Any
  \emph{open} specification containing region variables can be made
  closed by straightforward syntactic substitution with a (closed)
  \textit{region}.} specifications to sets of
$n$-dimensional indexing schemes. The interpretation is overloaded on
\emph{regions} in \Cref{fig:region-model} and on the top-level
of a specification \textit{spec} in \Cref{fig:spatial-model}.

@@

\noindent
The first four equations of \Cref{fig:region-model}
give the model of the region constants as sets of indexing
schemes. We illustrate with an example.

\begin{example}
For a $2$-dimensional stencil computation, then
\begin{align*}
& \interp{\stenCenS{2}{1}}_2 \\[-0.1em]
& \qquad = \; \{i\singleEntry{1}{2} \,|\, i \in \{\text{-}2,..., 2\}\!\setminus\!0 \}
  \cup  \{ \zeroEntry{1}{2} \,|\, \textit{true} \} \\[-0.2em]
& \qquad = \; \{-2\singleEntry{1}{2}, \, -\singleEntry{1}{2}, \,
  \singleEntry{1}{2}, \, 2\singleEntry{1}{2}\} \cup
  \{\zeroEntry{1}{2}\} \\[-0.2em]
& \qquad = \; \{\vtwoh{-2}{\infty}, \vtwoh{-1}{\infty}, \vtwoh{0}{\infty},
\vtwoh{1}{\infty}, \vtwoh{2}{\infty}\}
\end{align*}
Note the absorbing behaviour of $\infty$ with respect to
multiplication.
\end{example}
\noindent
The final two equations of \Cref{fig:region-model} give the models
of $\term{+}$ and $\term{*}$. For $\term{+}$ this is the
straightforward union of the models of subterms. The model of
$\term{*}$ takes the tensor $\otimes$ of subterm models. The tensor is
more involved; we give some informal intuition first.

@@

Regions are modelled as sets of index
schemes, that is
$\interp{\textit{region}}_n \in \mathcal{P}(\domainVal{}^n)$.
\Cref{fig:spatial-model} defines the top-level model of the
\textit{spec} syntax, which includes the multiplicity and
approximation modifiers. The domain of this interpretation
is $\textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$
where $\textsf{Mult}$ and $\textsf{Approx}$ are labelled variants
labelling our set-based model with additional information.

\begin{definition} \textsf{Mult}
and \textsf{Approx} are parametric labelled variant types
with injection function given by their definition:
%
\begin{align*}
\textsf{Mult} \, a & = \textsf{mult} \, a \,\mid\, \textsf{only} \, a \\
\textsf{Approx} \, a & = \textsf{exact} \, a \,\mid\, \textsf{lower} \, a \,\mid\,
\textsf{upper} \, a \,\mid\, \textsf{both} \, a \, a
\end{align*}
\eg{}, $\mathsf{lower}$ is an injection
$\mathsf{lower} : a \rightarrow \mathsf{Approx} \, a$ etc. The
$\textsf{Mult}$ type corresponds to the presence or absence of the
\term{readOnce} modifier as shown in \Cref{fig:spatial-model}.
The $\textsf{Approx}$ type corresponds to the presence or absence of
the spatial approximation modifier, with \textsf{exact} when there is
no such modifier and \textsf{lower} and \textsf{upper} for
\term{atLeast} and \term{atMost}. Later, it will be useful to model
pairs of lower and upper bounds provided by \textsf{both} but which is
not used
here.% These are used for labelled models, and other components later in.
\label{def:mult-and-approx}
\end{definition}

@@

The equality relation $\equiv$ is defined over \emph{region}
syntax which gives the following axioms:
%
\begin{itemize}
\item $\term{+}$ is idempotent, commutative, and associative;
\item $\term{*}$ is commutative, associative, and distributes over
$\term{+}$, \ie{},  $R \; \texttt{*} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{+} \; (R
                       \; \texttt{*} \; T)$.
Distributivity is
used later to give a normal form for specifications
akin to Disjunctive Normal Form (\cref{sec:inf-step4}).

@@

The equational theory provides not only an axiomatic semantics
for the language, which the user can use to rewrite their
specification, but it is also used extensively in the inference
procedure (\cref{subsec:inference}) to simplify specifications.

@@

\Cref{fig:inequations} defines a notion of sub-specifications
via the relation $<:^r$ on \textit{region}
syntax and then $<:$ on \textit{spec} syntax inductively.
Both relations are reflexive and transitive (we omit
these equations). The $<:^r$ relation is congruent
with respect to \term{+} and \term{*} $<:$ is congruent
with repect to \term{readOnce}, the equations of which we omit for brevity.

@@

\begin{theorem}[Approximation soundness]
Code consistent with a specification $S$
 is consistent with a more general specification $T$,
\ie{}, for all models $M$ of a piece of code:
\begin{equation*}
S <: T \; \wedge (\consA{M}{\interp{S}_n}) \;
\Rightarrow \; (\consA{M}{\interp{T}_n})
\end{equation*}
where $\consAName$ implements consistency in checking (\cref{subsec:checking}).
\end{theorem}

@@

\noindent
We outline here the procedures for checking conformance
of source code against specifications (\Cref{subsec:checking})
and for infering specifications from code (\Cref{subsec:inference}).
Both rely on a program analysis that converts array indexing
expressions into sets of index schemes. We outline this analysis
first (\cref{subsec:analysis}). Note that the analysis procedure
can be made arbitrarily more complex and wide-ranging independent
of the checking and inference procedures. At the moment, the analysis
is largely \emph{syntactic}, with only a small amount of
semantic interpretation of the code.

@@

The core analysis, captured by the function $\textit{analyse}$ below, then proceeds as follows.
For each assignment statement \textit{assg} within a loop whose left-hand side is an array
subscript on neighbourhood indices, a finite map is computed mapping
array variables to \emph{multisets} of indexing schemes:
%
\begin{align*}
\hspace{-0.5em}
\begin{array}{lr}
\textit{analyse}(a(\overline{e_1}) = e_2)
 := \qquad\qquad\qquad & \textit{where} \; \neigh(\overline{e_1}) \\[0.3em]
\multicolumn{2}{l}{\qquad \bigcup\{a' \mapsto \{
\textsf{schematic}(\bar{e})\} \mid a'(\bar{e}) \leftarrow \mathsf{flowsTo}(e_2)\}}
\end{array}
\end{align*}
%
The predicate $\neigh(\bar{e_1})$ classifies indices where each
$e \in \bar{e_1}$ is either a
neighbourhood index (based on
the induction variables of the loop, implicit here)
or a constant index.\footnote{If all indices $e \in \bar{e}$ are constants, then
  $\neigh(\bar{e})$ does not hold. Instead $a(\bar{e}$) is
  considered a constant by our analysis, rather
than an array subscript.}% contributing to a stencil.}
Thus, we focus on assignments to an array subscript where the LHS
indices $\overline{e_1}$ are neighbourhood indidces.  For
all array subscripts that flow to the right-hand side of this
statement, a finite map is constructed, mapping each array variable
to a multiset of indexing schemes for its subscripts, computed
with \textsf{schematic} (\Cref{def:schematic}).
Note \textsf{schematic} is undefined ($\bot$) if
$\bar{e}$ contains relative indices which are not neighbourhood
indices. The \textit{analysis} function is used by the
\textit{modelise} function which filters out those arrays
which contain undefined indexing schemes and adds multiplicity
information:
%
\begin{align*}
& \mathit{modelise} (\textit{assg})
= \textit{let} \;\; M = \mathit{analyse}(\textit{assg}) (a) \;\; \textit{in}
\\
& a \mapsto
\begin{cases}
\bot & \bot \in M \\
\textsf{only}(|M|) & \forall u \in M . \, \#(M, u) = 1 \\
\textsf{mult}(|M|) & \forall u \in M . \, \#(M, u) = n
\end{cases}
\end{align*}
where $\#(M, u)$ denotes the multiplicity of an element $u$ of the
multiset $M$ and $|M|$ denotes the set of elements in the multiset
$M$.
Thus, the analysis maps source code to index schemes, which is then
augmented with multiplicity information by $\mathit{modelise}$,
\ie{} values in $\mathsf{Mult}(\mathcal{P}((\domainVal{})^n))$.

@@

Checking verifies the access pattern of a stencil computation in the
source language against its associated specifications. The model of
(\Cref{sec:semantics}) guides this process.  Checking
proceeds by generating a model from a specification and generating
a model from the source code, comparing them for
\emph{consistency} base on index scheme
\emph{generalisation} (\Cref{def:generalisation}).

@@

A stencil kernel
 $v(\overline{e_1}) = e_2$
conforms to $S$ if the predicate $\textbf{check}_n(S, v(\overline{e_1}) = e_2, a)$ holds,
defined:
\begin{align*}
& \textbf{check}_n(S, \textit{assg}, a) =
  \consSub{n}{\textit{modelise}(\textit{assg})(a)}{\interp{S}_n}
&
\end{align*}
Thus, checking applies \textit{modelise}
(\Cref{subsec:analysis}) to extract index schemes from source code,
augmented with multiplicitly for the whole set of schemes.
This is compared with the model of the specification $\interp{S}_n$ by
the consistency predicate $\consAName$.
\end{definition}

@@

\begin{definition}
Consistency relates models of source code to models
of specifications (which are bounded) via predicates:
\begin{align*}
\consName_n & \subseteq \mathcal{P}(\domainVal{}^n)
\times \textsf{Approx}(\mathcal{P}(\domainVal{}^n)) \\
\consAName_n & \subseteq \textsf{Mult}(\mathcal{P}(\domainVal{}^n))
\times \textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))
\end{align*}
%
where $\consName$ relates models without multiplicity.
\Cref{fig:consistency} defines the two predicates, where for
brevity we make implicit the subscript $n$ defining the
dimensionality.  We denote sets of indexing schemes
$\mathcal{P}((\domainVal{})^n)$
from source code model by $M$ and from a specification model by $M$.

@@

To aid adoption of our approach to legacy code, we provide
an inference procedure for generating specifications from code, which
are inserted automatically as comments.
We illustrate inference using the
 five-point stencil of
 \Cref{exm:checking}.

@@

The first step performs the
analysis of \cref{subsec:analysis} to produce a
model of the code in terms of indexing schemes.
The \emph{analyse} and \emph{modelise} procedures before yielded the finite map:
%
\begin{align*}
\texttt{a} \mapsto \mathsf{only}(\{\vtwoh{-1}{0}, \vtwoh{1}{0},
          \vtwoh{0}{-1}, \vtwoh{0}{1}, \vtwoh{0}{0}\})
\end{align*}
%

@@

\noindent
Each indexing scheme $u \in U(a)$ is converted to a \emph{unit} span
$u \mapsto [u \; u]$, defining the lower and upper bound of an
$n$-dimensional cube of side length $1$. For the five-point stencil
example there are five unit spans, illustrated in
\Cref{fig:inference-steps-informal}(a).

@@

\begin{definition}\label{def:span-coalesc}
  Span coalescing $\bullet$ is partial, defined for spans contiguous in the
  first dimension:
%%
\begin{align*}
\vect{x} \bullet \vect{y}
= (\vect{x}^L, \vect{y}^U) \;\;\;\, \textit{if} \;\;\; \vect{x}^U_1 + 1 = \vect{y}^L_1 \; \wedge \;
\vect{x}_{2:n} = \vect{y}_{2:n} %\\
%\bot  & \textit{otherwise}
%\end{cases}
% $m$-dimensions into coalesced spans which vary in
% $m+1$-dimensions.
\end{align*}
%%
\end{definition}
\noindent
To coalesce regions which are contiguous in other dimensions
other than the first, we generate all rotations of the spans in
$n$-dimensional space by permutation on spans, and fold together contiguous
regions using $\bullet$ (above). This
generates all possible ways of coalescing regions. For our example,
the unit spans in \Cref{fig:inference-steps-informal}(a)
are coalesced into the contiguous spans in
\Cref{fig:inference-steps-informal}(b).
This procedure is iterated until a fixed point is reached (in one step
in this example).
 Then any spans that are completely contained by any other span are deleted,
leaving a minimal set of spans (which may overlap, but none of which
fully contains another), \Cref{fig:inference-steps-informal}(c).
The final spans are used to synthesise concrete stencil specification
syntax (\cref{sec:inf-step4}), where
each span becomes a product \texttt{*} of regions which are
combined by the sum \texttt{+}.

@@

\begin{definition} The $\mathsf{toSpec}$ constructor is defined
in \Cref{fig:to-spec}, where the lower and upper bounds
determine region constants; $\mathsf{toSpec} \, d \, (l, u)$ is always defined
 if $l$ less than or equal to $u$.

@@

\paragraph{Hypothesis 1: Stencil computations
(as in \Cref{def:stencil}) are a common programming pattern
in numerical code}
